---
title: "STAT 331/531: Final Project"
author: "Kylie Scharf, Emma Durler, Jordan, Mason"
format: 
  html:
    embed-resources: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

```{r}
#| label: setup
#| output: false
library(tidyverse)
babies <- read_csv(here::here("children_per_woman_total_fertility.csv"))
life <- read_csv(here::here("lex.csv"))

```

```{r}
#| output: false
#data cleaning
babies <- babies |>
  select(country, `1900`:`2000`) |>
  pivot_longer(`1900`:`2000`,
               names_to = "Year",
               values_to = "Babies_Per_Woman")

life <- life |>
  select(country, `1900`:`2000`) |>
  pivot_longer(`1900`:`2000`,
               names_to = "Year",
               values_to = "Life_Expectancy")

full <- inner_join(babies, life)

```


```{r}
#| output: false
#Linear Regression

full |>
  mutate(decade = case_when(Year %in% c(1900:1910) ~ "00s",
                            Year %in% c(1910:1920) ~ "10s",
                            Year %in% c(1920:1930) ~ "20s",
                            Year %in% c(1930:1940) ~ "30s",
                            Year %in% c(1940:1950) ~ "40s",
                            Year %in% c(1950:1960) ~ "50s",
                            Year %in% c(1960:1970) ~ "60s",
                            Year %in% c(1970:1980) ~ "70s",
                            Year %in% c(1980:1990) ~ "80s",
                            Year %in% c(1990:2000) ~ "90s"
                            )
         ) 
```

# Description of the data

We wanted to investigate the potential relationship between babies per woman and life expectancy in different countries throughout the world over time. We got our data from the Gap Minder website and pivoted and joined our data sets (children_per_woman_total_fertility.csv and lex.csv) to include only 5 necessary variables: Country, Year, Babies_Per_Woman, and Life_Expectancy, and Decade.

